---
title: "dplyr for NFL Data"
subtitle: "Data Cleaning and Transformation for the NFL Project"
author: "Erin Mansfield"
date: last-modified
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
    code-tools: true
    code-summary: "Show the code"
    highlight-style: atom-one
execute:
  echo: true
  warning: false
  message: false
---

# Getting Started

This tutorial shows that I used dplyr to clean, transform, and prepare the NFL game level dataset fro the Data Storytelling Project. 
-Clean game dates
-Extract week numbers
-Create scoring variables 
-Filter to regular season games
-Summarize scoring by season and week 

<br>

# Setup

## Install (one-time)

```{r}
# install.packages("tidyverse")

```

## Load packages

```{r}
library(tidyverse)
library(lubridate)
library(stringr)
```

<br>

# A Quick Look at a Dataset

We will use the built-in dataset **`mpg`**.

```{r}
scores |> glimpse()
```

<br>

# The Pipe Operator `|>`

The pipe sends the output of one step into the next step.

```{r}
scores |>
  select(Seasons, Week, HomeTeam, AwayTeam, HomeScore, AwayScore)
```

This is equivalent to:

```{r}
select(scores, Season, Week, HomeTeam, AwayTeam, HomeScore, AwayScore)
```

✅ Use pipes to write code in a clear, step-by-step style.

<br>

# `select()` (Choose Columns)

## Select a few columns

```{r}
scores |>
  select(Season, Week, GameDate, HomeTeam, AwayTeam)
```

## Select a range of columns

```{r}
scores |>
  select(Season, Week, GameDate)
```

## Remove columns

```{r}
scores |>
  select(-Stadium, -Network, -Referee)
```

<br>

# `filter()` (Choose Rows)

## Filter by a condition

```{r}
scores |>
  filter(GameStatus == "final", str_detect(Week, "^WEEK"))
```

## Multiple conditions with AND

```{r}
scores |>
  filter(Season == 2024)
```

## OR conditions using `%in%`

```{r}
scores |>
  mutate(total_points = HomeScore + AwayScore) |>
  filter(total_points > 50)
```

<br>

# `arrange()` (Sort Rows)

## Sort ascending

```{r}
scores |>
  mutate(total_points = HomeScore + AwayScore) |>
  arrange(total_points)
```

## Sort descending

```{r}
scores |>
  mutate(total_points = HomeScore + AwayScore) |>
  arrange(desc(total_points))
```


<br>

# `mutate()` (Create New Variables)

## Create total points and point differential

```{r}
scores |>
  mutate(
    total_points = HomeScore + AwayScore, 
    point_diff = abs(HomeScore - AwayScore)
  )
```

## Extract numeric week number

```{r}

scores |>
  mutate(week_num = as.numeric(str_extract(Week, "\\d+")))

```

## Clean game dates

```{r}
scores |>
  mutate(
    GameDate_clean = str_remove(GameDate, "(st|nd|rd|th)$"),
    GameDate_full  = paste(GameDate_clean, Season),
    GameDate       = mdy(GameDate_full)
  )
```

<br>

# `summarise()` (Compute Summary Statistics)

## Average Points per Season

```{r}
scores_clean |>
  group_by(Season) |>
  summarise(
    avg_points = mean(total_points, na.rm = TRUE),
    .groups = "drop"
  )

```

## Highest scoring week in each season

```{r}
scores_clean |>
  group_by(Season, week_num) |>
  summarise(
    avg_points = mean(total_points, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(Season, desc(avg_points))

```

✅ Tip: Use `na.rm = TRUE` if your data has missing values.

<br>

# `group_by()` + `summarise()` (Grouped Summaries)

## Average point differential by season

```{r}
scores_clean |>
  group_by(Season) |>
  summarise(
    mean_diff = mean(point_diff),
    .groups = "drop"
  )

```

## Number of games per week

```{r}
scores_clean |>
  group_by(week_num) |>
  summarise(n_games = n(), .groups = "drop")

```

<br>

# `count()` (Fast Frequency Tables)

```{r}
scores_clean |> count(Season)

```

Count games by home team

```{r}
scores_clean |> count(HomeTeam, sort = TRUE)

```

<br>

# `distinct()` (Unique Values)

```{r}
scores_clean |> distinct(HomeTeam)
```

<br>

# `slice_*()` (Pick Specific Rows)

## Top rows- 10

```{r}
scores_clean |>
  arrange(desc(total_points)) |>
  slice_head(n = 10)
```


<br>

# `across()` (Apply Functions to Many Columns)

Example: compute means for multiple numeric columns.

```{r}
scores_clean |>
  summarise(
    across(c(HomeScore, AwayScore), mean)
  )

```


<br>

# Joining Data (Combining Tables)

Joins combine tables using a matching key column.

We will create two small example tables.

```{r}
scores_clean |>
  left_join(team_info, by = "HomeTeam")

```


<br>

# A Typical dplyr for NFL Example



```{r}
scores |>
  filter(GameStatus == "FINAL", str_detect(Week, "^WEEK")) |>
  mutate(
    total_points = HomeScore + AwayScore,
    week_num = as.numeric(str_extract(Week, "\\d+"))
  ) |>
  group_by(Season) |>
  summarise(
    avg_points = mean(total_points),
    n_games = n(),
    .groups = "drop"
  ) |>
  arrange(desc(avg_points))
```

<br>

# Practice Problems ✅

1. Select only Season, Week, HomeTeam, AwayTeam, and scores
2. Filter to games where total points exceed 60
3. Sort games by point differential
4. Create a variable home_win indicating whether the home team won
5. Compute the average point differential by season
6. Find the top 5 highest scoring seasons

