---
title: "ggplot2 for NFL Data"
subtitle: "Visualization of Scoring Trends in the NFL"
author: "Erin Mansfield"
date: last-modified
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
    code-tools: true
    code-summary: "Show the code"
    highlight-style: atom-one
execute:
  echo: true
  warning: false
  message: false
---

# Getting Started

This page explains how I used ggplot2 to visualize NFL scoring patterns for the Data Storytelling Project. All examples used the cleaned dataset scores_clean, which includes: 
-total points, point differential, week number, season, home/away scores, and cleaned date scores

We learned how to: 
-create scatterplots, line charts, and bar charts using NFL data
- Map scoring variables
-Compare season and weeks
-Customize themes, colors, and labels
-Graphics for analysis and storytelling

<br>

# Setup

## Install (one-time)

```{r}
# install.packages("tidyverse")
```

## Load packages

```{r}
library(tidyverse)
library(lubridate)
library(stringr)
```

<br>

# The ggplot2 Grammar of Graphics

Most ggplot charts follow this structure:

```{r}
ggplot(scores_clean, aes(x = X_VAR, y = Y_VAR)) +
  geom_...() +
  labs(title = "...", x = "...", y = "...") +
  theme_minimal()
```

Key pieces:

- **Data**: NFL Dataset
- **Aesthetics (`aes`)**: how variables map to visual elements
- **Geoms**: the type of plot (points, lines, bars, etc.)
- **Labels**: title, subtitle, axes, caption
- **Theme**: overall style

<br>

# Example Dataset

We will use the built-in dataset `mpg` (fuel economy data).

```{r}
scores_clean |> glimpse()
```

<br>

# Scatterplots

## Total points vs point differential

```{r}
ggplot(scores_clean, aes(x = point_diff, y = total_points)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  labs(
    title = "Total Points vs Point Differential",
    x = "Point Differential",
    y = "Total Points"
  ) +
  theme_minimal()
```

## Color by Season

```{r}
ggplot(scores_clean, aes(x = point_diff, y = total_points, color = Season)) +
  geom_point(alpha = 0.6) +
  theme_minimal()
```


✅ Rule of thumb:  
- `color = ...` inside `aes()` means **it changes by data values**  
- `color = "blue"` outside `aes()` means **fixed color**

<br>

# Trend Lines

## Smooth trend of scoring over time

```{r}
ggplot(scores_clean, aes(x = Season, y = total_points)) +
  geom_point(alpha = 0.3) +
  geom_smooth(se = FALSE, color = "darkred") +
  labs(
    title = "Trend in NFL Scoring (2010–2025)",
    x = "Season",
    y = "Total Points"
  ) +
  theme_minimal()
```



<br>

# Faceting (Small Multiples)

Compare scoring patterns by season

```{r}
ggplot(scores_clean, aes(x = week_num, y = total_points)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  facet_wrap(~ Season) +
  labs(
    title = "Weekly Scoring Patterns by Season",
    x = "Week",
    y = "Total Points"
  ) +
  theme_minimal()
```

<br>

# Bar Charts

Bar charts are usually used for **categorical variables**.

## Number of games per season

```{r}
ggplot(scores_clean, aes(x = factor(Season))) +
  geom_bar(fill = "darkgreen") +
  labs(
    title = "Number of Games per Season",
    x = "Season",
    y = "Count"
  ) +
  theme_minimal()
```

## Average points per week

```{r}
week_summary <- scores_clean |>
  group_by(week_num) |>
  summarise(avg_points = mean(total_points), .groups = "drop")

ggplot(week_summary, aes(x = week_num, y = avg_points)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Average Points by Week (2010–2025)",
    x = "Week",
    y = "Average Points"
  ) +
  theme_minimal()
```

<br>


# Line Charts

Line charts require an x-variable that has a meaningful order (often time).

We will build a small example dataset.

Average points per season

```{r}
season_summary <- scores_clean |>
  group_by(Season) |>
  summarise(avg_points = mean(total_points), .groups = "drop")

ggplot(season_summary, aes(x = Season, y = avg_points)) +
  geom_line(color = "darkblue", linewidth = 1.2) +
  geom_point(size = 2, color = "darkblue") +
  labs(
    title = "Average Total Points per Season",
    x = "Season",
    y = "Average Points"
  ) +
  theme_minimal()

```


<br>

# Labels and Titles

Use `labs()` to make your charts understandable.

```{r}
ggplot(scores_clean, aes(x = week_num, y = total_points)) +
  geom_point(alpha = 0.4) +
  labs(
    title = "NFL Scoring by Week",
    subtitle = "Regular-season games only",
    x = "Week Number",
    y = "Total Points",
    caption = "Source: NFL Game Logs (2010–2025)"
  ) +
  theme_minimal()
```

<br>

# Themes

Themes change the overall style.

```{r}
ggplot(scores_clean, aes(x = Season, y = total_points)) +
  geom_point(alpha = 0.4) +
  theme_classic()
```

Try a few common ones:

```{r}
theme_light()
theme_minimal()
theme_bw()
```

<br>

# Changing Colors (Manually)

If you want a **custom palette**, you can set:

```{r}
ggplot(scores_clean, aes(x = Season, y = total_points, color = factor(Season))) +
  geom_point(alpha = 0.6) +
  scale_color_viridis_d() +
  theme_minimal()
```


<br>

# Saving Plots

Use `ggsave()` to export a plot.

```{r}
p <- ggplot(season_summary, aes(x = Season, y = avg_points)) +
  geom_line() +
  theme_minimal()

ggsave("season_scoring_trend.png", plot = p, width = 8, height = 5, dpi = 300)
```

<br>

# Practice Problems ✅

1. Create a scatterplot of week_num vs total_points
2. Color points by Season and set alpha = 0.5
3. Add a smooth trend line 
4. Use facet_wrap to compare scoring across seasons

<br>

